---
title: "Covid !9 Analysis"
author: "Group Work"
date: "2023-11-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

```



```{r, message=FALSE}

# Load required libraries
library(tidyverse)
library(lares) # show the structure of data
library(vtable) # show the mean, sd, min, max
library(caret) 
library(psycho) # Standardizing
library('DescTools')
library(ggcorrplot)
library(plotly) # Using plotly for interactive plots
library(fastDummies) # for dummy encoding

```



## About the Dataset

The data is about Coronavirus (COVID-19) Vaccinations

According to the website, the vaccination dataset uses the most recent official numbers from governments and health ministries worldwide. Population estimates for per-capita metrics are based on the United Nations World Population Prospects. Income groups are based on the World Bank classification. Source: https://ourworldindata.org/covid-vaccinations


**NB: The dataset should have at least 7000 rows and 10 columns after cleaning and there is not any upper
bound. The objectives of the DEP project are based on the domain knowledge of data. Need to complete
the following tasks during the development of this project.**


## Loading the Dataset


```{r}

# loading the data after downloading
data <- read.csv('owid-covid-data.csv')

dim(data)

# overview of the data
head(data, 5)

str(data)

```



















